Linear analysis of the vertical shear instability in astrophysical disks

### 2015-02-07 linear_iso.tex
updating
undergoing major revisions

\section{Analytical discussion of vertically isothermal disks}\label{analytical}
We begin with a discussion of vertically isothermal disks
where $\Gamma=1$, then $c_s$ is a constant in the shearing box. We
first examine the limiting case of isothermal perturbations
($\gamma\equiv1$ and any value of $\beta$, or any value of $\gamma$ with
$\beta = 0$) before considering the effect of thermal relaxation. In the
final section we consider adiabatic disks ($\gamma>1$ and
$\beta\to\infty$) to demonstrate an important limitation of the 
local model.

\subsection{Approximate governing equation for general perturbations}\label{approx_gov}
To facilitate a simple discussion, we derive a governing
equation with the following approximations:

\begin{enumerate}
\item We consider eigenvalues such that
  $|\sigma^2|\ll \kappa^2$. Then we can replace $D=\kappa^2 -\sigma^2\to
  \kappa^2$. % The problem simplifies because the eigenvalue now only
  % appears in the last term in Eq. \ref{iso_ode}.
  We call this the
  low-frequency approximation. 
\item We are mainly interested in the effect of vertical shear, i.e. $q\neq
  0$. This is already captured explicitly in the linearized equations through
  the term $d\Omega/dz$. We therefore simplify
  further by ignoring the vertical dependence of $\Omega$ and $\kappa$
  elsewhere. For clarity, we replace $\kappa^2$ with $\Omega_k^2$.

@@

It is convenient to introduce the dimensionless variables
\begin{align}
  \hat{z} = z/\Hiso,\quad \hat{k}=k_x\Hiso, \quad \hat{\sigma} = \sigma/\Omega_k,
\end{align}
then Eq. \ref{vertiso_gov} becomes
\begin{align}
  0=& \delta v_z ^{\prime\prime} + \left(1 + \ii\epsilon
    q\hat{k}\right)\ln\rho^{\prime}\delta v_z^\prime \notag\\
  &+
  \left[\hat{\sigma}^2\left(\hat{k}^2+\chi\right) +
    \left(  \chi + \ii \epsilon q\hat{k}\right)\ln\rho^{\prime\prime}\right.
    \notag\\
    &\left.- \ln\rho^{\prime
      2}\left(\khat^2 -
      \ii\epsilon
      q\hat{k}\right)\left(1 - \chi\right)\right]\delta v_z,\label{vertiso_gov_nondim}
\end{align}
where $^\prime$ denotes $d/d\zhat$. Eq. \ref{vertiso_gov_nondim} will
be the basis for our discussion of vertically isothermal disks,
although further simplifications or maniputations will be performed
depending the type of perturbations being considered.

@@

\subsubsection{Stability in the absence of vertical shear}
When $q=0$, Eq. \ref{iso_ode3} is
\begin{align}\label{hermite_ode}
  \left[w(\hat{z})W^\prime \right]^\prime + nW
  w(\hat{z}) =0, 
\end{align}
where we have defined a new eigenvalue
\begin{align}
  n \equiv \hat{\sigma}^2(1+\hat{k}^2). 
\end{align} 
Eq. \ref{iso_ode3} is Hermite's differential equation. If we impose
that the kinetic energy density remain bounded at infinity, then  
\begin{align}
  W \propto \He_n(\hat{z}),
\end{align}
and $n$ is a non-negative integer. This translates to a real
eigenfrequency $\sigma$ such that
\begin{align}
  \left|\hat{\sigma}\right| = \sqrt{n}
  \left(1+\hat{k}^2\right)^{-1/2}. 
\end{align}
Since we have assumed $|\sigma^2|\ll \kappa^2\sim \Omega_k^2$, our
analysis is only valid for large wavenumbers such that $\hat{k}^2\gg   
n-1$. Physically, this corresponds to radial length-scales much
smaller than the local disk scale height.

@@

In Appendix \ref{pert_theory} we give an alternative method to
infer instability in the presence of vertical shear without explicitly
solving the governing equation.

>>>>

\section{Analytical discussion of vertically isothermal
  disks in the framework of the vertically global shearing
  box}\label{analytical}  
We begin with a discussion of vertically isothermal disks
where $\Gamma=1$, then $c_s$ is a constant. We  
first examine isothermal perturbations ($\gamma\equiv1$ and any value
of $\beta$, or any value of $\gamma$ with $\beta = 0$) before
considering the effect of finite thermal relaxation. 
In order to facilitate an analytical discussion, we introduce a number
of approximations below that allows a simple governing equation to be
derived and directly compared with previous works.

\subsection{The linear problem in the vertically global
  shearing box} 
We connect the above linearized equations to the vertically global
shearing box (VGSB) formalism developed by \cite{mcnally14}. The VGSB
is a generalization of the standard shearing box \citep{goldreich65}
by allowing the shear flow in the shearing box to be height-dependent.

The linear problem in the VGSB framework can be derived from 
Eq. \ref{lin_mass}---\ref{lin_vz} by setting $\hat{g}_c=0$, since 
the VGSB, like the standard shearing box, is homogeneous in the radial 
direction in equilibrium. The resulting linear problem is very similar
to that for local axisymmetric waves in accretion disks already
discussed by \cite{lubow93}, but modified by the presence of vertical
shear and thermal relaxation. This is convenient as it allows us to
directly apply their solution methods.

% but modified by the vertical shear term
% $r\p_z\Omega\delta \delta v_z$.

After setting $\hat{g}_c=0$ in Eq. \ref{lin_mass}, \ref{lin_vx} and
\ref{lin_energy}, we can eliminate the perturbed velocities between
the linearized equations to obtain a pair or ordinary differential
equations for $W,Q$ as
\begin{align}
  &\frac{\sigma^2}{c_s^2}Q + \frac{\sigma^2k_x^2}{D}W \notag\\ 
  &=\left(\frac{\ii
      k_x r}{D}\frac{d\Omega^2}{dz} -
    \frac{d\ln{\rho}}{dz}\right)\left[\frac{dW}{dz} +
    \frac{d\ln{\rho}}{dz}\left(W-Q\right)\right] \notag\\
  &- \frac{d^2W}{dz^2} - \frac{d\ln{\rho}}{dz}\left(\frac{dW}{dz} -
    \frac{dQ}{dz}\right) - \frac{d^2\ln{\rho}}{dz^2}\left(W-Q\right),\label{ode_w}\\
  &\sigma^2W - \frac{\gamma}{\Gamma}\sigma^2 Q +
  \frac{\ii\sigma}{t_c}\left(W-\frac{Q}{\Gamma}\right)\notag\\
  &=c_s^2\frac{d\ln{\rho}}{dz}\left(\frac{\gamma}{\Gamma} - 1\right) 
  \left[\frac{dW}{dz} + \frac{d\ln{\rho}}{dz}\left(W-Q\right)\right],\label{ode_Q} 
\end{align}
where
\begin{align}
  D \equiv \kappa^2 - \sigma^2,
\end{align}
and we have replaced the notation $\p_z$ by $d/dz$ since we are
considering solutions at a fixed radius.

The VGSB is a self-consistent framework but extrapolating results 
obtained from the VGSB to the global disk has limitations. This is
because within the VGSB framework, $d\Omega/dz$ can be freely
specified. In effect, by adopting the VGSB framework, we are studying
dynamics in the standard shearing box subject to an additional
azimuthal force $\delta v_zr d \Omega/dz$ with $d\Omega/dz$ as an
input parameter. This is different to the global problem because in
that case, the vertical shear is a result of the background radial
disk structure, which is ignored in the VGSB.

We further discuss caveats of the VGSB in Appendix \ref{global_corr}.
However, we find that for perturbations with small radial lengthscales
(large $|k_x|$) and fast thermal relaxation ($\beta\ll1$), which
characterizes the VSI, the VGSB framework is adequate for the purpose
of a simple analytical discussion.

% In effect, we are studying dynamics in the standard shearing box
% subject to an additional azimuthal force $-v_zr_0 d \Omega_0/dz$.

% $d\Omega_0/dz$ is a free parameter (as are $\Omega_0$ and $S_0$). 
% This is different to the global problem because in that case, the
% vertical shear is a result of the disk radial structure, which is
% ignored in the VGSB equilibria. 
% In effect, we are studying dynamics in the standard shearing box
% subject to an additional azimuthal force $-v_zr_0 d \Omega_0/dz$. This
% force is inspired by the vertical shear in the global disk, and its
% functional form is taken from global equilibrium.

% \subsection{Limitations}
% Before using the above equations to study the VSI, it is very important to be
% aware of several limitations of the VGSB model. The VGSB momentum  
% equations have the same form as that for the standard shearing box 
% \citep{goldreich65}, except for the appearence of the term 
% $v_zr_0d\Omega_0/dz$. This term appears in the global stability
% problem and is the only term associated with vertical shear that is 
% ported from the global disk into the local model.

% In Appendix \ref{global_corr} we make the connection between the
% linear problem in the VGSB to that in the global disk model. We
% emphasize that not all calculations in the local model can be 
% consistently interpreted in the global context. However, for
% perturbations with small radial lengthscales (large $|k_x|$) and fast
% thermal relaxation ($\beta\ll1$), which characterizes the VSI, the
% VGSB framework is adequate.

\subsection{Approximate governing equation}\label{approx_gov}
We now derive a governing equation with further approximations:
\begin{enumerate}
\item We consider eigenvalues such that
  $|\sigma^2|\ll \kappa^2$. Then we can replace $D=\kappa^2 -\sigma^2\to
  \kappa^2$. % The problem simplifies because the eigenvalue now only
  % appears in the last term in Eq. \ref{iso_ode}.
  We call this the
  low-frequency approximation. 
\item We are mainly interested in the effect of vertical shear, i.e. $q\neq
  0$. This is already captured explicitly in the linearized equations through
  the term $d\Omega/dz$. We therefore simplify
  further by ignoring the vertical dependence of $\Omega$ and $\kappa$
  elsewhere. For clarity, we replace $\kappa^2$ with $\Omega_k^2$.

@@

It is convenient to introduce the dimensionless variables
\begin{align}
  \hat{z} = z/\Hiso,\quad \hat{k}=k_x\Hiso, \quad \hat{\sigma} = \sigma/\Omega_k,
\end{align}
then Eq. \ref{vertiso_gov} becomes
\begin{align}
  0=& \delta v_z ^{\prime\prime} + \left(1 + \ii\epsilon
    q\hat{k}\right)\ln\rho^{\prime}\delta v_z^\prime \notag\\
  &+
  \left[\hat{\sigma}^2\left(\hat{k}^2+\chi\right) +
    \left(  \chi + \ii \epsilon q\hat{k}\right)\ln\rho^{\prime\prime}\right.
  \notag\\
  &\left.- \ln\rho^{\prime
      2}\left(\khat^2 -
      \ii\epsilon
      q\hat{k}\right)\left(1 - \chi\right)\right]\delta v_z,\label{vertiso_gov_nondim}
\end{align}
where $^\prime$ denotes $d/d\zhat$. Eq. \ref{vertiso_gov_nondim} will
be the basis for our discussion of vertically isothermal disks. % ,
% although further simplifications or maniputations will be performed
% depending the type of perturbations being considered.

@@

\subsubsection{Stability in the absence of vertical shear}
When $q=0$, Eq. \ref{iso_ode3} is
\begin{align}\label{hermite_ode}
  \left[w(\hat{z})W^\prime \right]^\prime + nW
  w(\hat{z}) =0, 
\end{align}
where we have defined a new eigenvalue
\begin{align}
  n \equiv \hat{\sigma}^2(1+\hat{k}^2). 
\end{align} 
Eq. \ref{hermite_ode} is Hermite's differential equation. If we impose
that the kinetic energy density remain bounded at infinity, then  
\begin{align}
  W \propto \He_n(\hat{z}),
\end{align}
and $n$ is a non-negative integer. This translates to a real
eigenfrequency $\sigma$ such that
\begin{align}
  \left|\hat{\sigma}\right| = \sqrt{n}
  \left(1+\hat{k}^2\right)^{-1/2}. 
\end{align}
Since we have assumed $|\sigma^2|\ll \kappa^2\sim \Omega_k^2$, our
analysis is only valid for large wavenumbers such that $\hat{k}^2\gg   
n-1$. Physically, this corresponds to radial length-scales much
smaller than the local disk scale height.

@@

% In Appendix \ref{pert_theory} we give an alternative method to
% infer instability in the presence of vertical shear without explicitly
% solving the governing equation.

### 2015-02-07 numerical.tex
undergoing major revisions

\section{Numerical solutions}
In this section we numerically solve the linear stability problem
corresponding to Eq. \ref{lin_mass}---\ref{lin_energy}. 
We relax the thin-disk, low-frequency and nearly-Keplerian
approximations used in the discussion above, and 
use the full, non-Keplerian expressions of $\rho$, $\Omega$ and
$\kappa$.
%, so that we may consistently consider vertical domains up to
%$z\sim r$.

@@

%mention mode selection -> focus on fundamental corrugation ->
%simulations eventually dominated by them 
%choice of kx is O(10), since nelson observe kxH \sim 30

\subsection{VSI in a locally isothermal, neutrally stratified disk}\label{vertiso_pertiso} 
%  disk model is a nearly-vertically
% isothermal disk with $\Gamma=1.011$, $p=-1.5$, $q=-1$,
% $\epsilon=0.1$. 
We first calculate the VSI in a nearly vertcally isothermal, neutrally
straitified disk ($N_z^2\equiv0$) by setting
$\gamma=\Gamma=1.011$ and $\beta=10^{-3}$. Other disk parameters are 
$(p,q,\epsilon)=(-1.5,-1,0.1)$ and the vertical domain is $\zmax =
10\epsilon r$. This setup is very similar to that adopted in
\cite{mcnally14}, who calculated the VSI in vertically isothermal
disks subject to isothermal perturbations ($\Gamma=\gamma=1$).  This
allows us to test our numerical code.

@@

Fig. \ref{iso_eigen_kx} compares the numerically-obtained eigenvalues
for the fundamental mode and that caluclated from
Eq. \ref{simple_growth} (with $L=1$). For completeness we
included cases with $\khat\ll 1$ even though the local
approximation does not hold in this regime. The agreement is good for
$\khat\gtrsim10$. The mismatch is significant for $\khat\ll1$ since
the low-frequency approximation used to obtain Eq. \ref{simple_growth}
is no longer valid: $|\omega|\sim\Omega_k$ as $\khat\to 0$. This means
that our analytic results should only be applied to perturbations with
$\khat\gg 1$, i.e. small radial-lengthscales.

@@

%particular kx 
As an example of an unstable mode, we consider a perturbation
wavenumber $\khat=20\pi$ \citep[or $k_x = 200\pi/r$ as used
in][]{mcnally14}. Fig. \ref{lowfreq_eigen} shows the eigenvalues
$\sigma = \omega + \ii\nu$. The fundamental VSI mode, with one node in $W$, has
the smallest $|\sigma|$ and is shown in Fig. \ref{lowfreq_eigenfunc}. 
The eigenfunction $W$ is qualitatively proportional to 
$z$ (especially for $z\to0$),  as expected for the fundamental mode
according to explicit solutions developed in \S\ref{iso_poly}.    
These plots are consistent with those in \cite{mcnally14}.

The growth rate of the fundamental VSI calculated from
Eq. \ref{simple_growth} with $L=1$ and the numerically-obtained value
are  
\begin{align*}
  \nu = 0.2606\epsilon\Omega_k &\quad \text{(from
    Eq. \ref{simple_growth})},\\
  \nu = 0.2575 \epsilon \Omega_k &\quad \text{(numerical)}.
\end{align*}
This agreement is surpsingly good, given that
Eq. \ref{simple_growth} assumes a thin, nearly-Keplerian disk in the 
low-frequency approximation, and 
imposes a different boundary condition to that in the numerical
calculation. These values are also close to that found by
\cite{mcnally14}. The eigenfrequency is not sensitive to the
vertical boundary condition because the expression for $\sigma$, given
by Eq. \ref{integral_relation1}, involve integrands with $\rho$ as the  
weight function, which rapidly decays away from the midplane.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/eigenvalues_iso}
  \caption{Eigenvalues of the VSI in a neutrally-stratified,
    nearly-vertically isothermal disk  ($\gamma=\Gamma=1.011$) with 
    $(p,q,\epsilon)=(-1.5,-1,0.1)$ and $\beta=10^{-3}$. The
    perturbation wavenumber is $\khat=20\pi$. \label{lowfreq_eigen} 
  }
\end{figure}

\begin{figure}
  \includegraphics[width=\linewidth]{figures/eigenvector_iso}
  \caption{Eigenfunction of the fundamental VSI,
    corresponding to the bottom-left eigenvalue displayed in
    Fig. \ref{lowfreq_eigen} (smallest $|\sigma|$). The real
    (imaginary) parts of $W$ are plotted as solid (dotted) lines. 
    \label{lowfreq_eigenfunc}
  }
\end{figure}

@@

Fig. \ref{relax_growth_num} shows the growth rates of the fundamental  
mode as a function of $\khat$. This plot is largely consistent with  
Fig. \ref{relax_disp_fig} except for $\khat\lesssim 1$
where the low-frequency approximation fails. (In any case, the local model not valid
for $\khat\ll1$.) For $\khat\gtrsim 10$ introducing thermal
relaxation rapidly stabilizes the fundamental mode.

% For $\beta=0.063$ at $\khat\gtrsim50$ we found eigenfunctions had 
% large perturbations near the vertical boundaries, which may be 
% responsible for the non-smoothness in the curve at large $\khat$. 
%corrugation mode
% We first obtain the eigenvalues for the
% fundamental mode for $k_xH_\mathrm{iso}=0$, then vary parameters
% slowly to obtain eigenvalues for other values of $k_x$ or $\beta$.  
%likely attributable to the free surface boundary condition imposed in
%the numerical calculation (cf. regularity condition at infinity in
%\S\ref{disp_relax}).

Fig. \ref{relax_growth_num} also confirms our analytical discussion that
introducing a small but finite cooling time stabilizes the fundamental
VSI, but there is a preferable wavenumber that maximizes this effect (\S\ref{relax_pert}).
This occurs at $\khat= O(10)$ for the current disk model.

\begin{figure}
   \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm
   0cm]{figures/compare_eigen_imag_bloop} 
  % \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  % 1cm]{figures/compare_eigen_imag_bloop2} 
  \caption{Growth rates of the fundamental VSI mode in 
    nearly-vertically isothermal disks ($\Gamma=1.011$) with
    $(p,q,\epsilon)=(-1.5,-1,0.05)$, evolved with $\gamma=1.4$ under a
    range of thermal relaxation timescales   
    $\beta$. The eigenfrequencies are calculated by numerically 
    solving the full linear eigenvalue problem. \label{relax_growth_num}}   
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/eigenvectorW_beta0d01} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/eigenvectorW_beta0d125} 
  \caption{Numerically-calculated eigenfunction of the fundamental VSI
    mode in a nearly vertically isothermal disk ($\Gamma=1.011$) with
    $(p,q,\epsilon)=(-1.5,-1,0.05)$, evolved with $\gamma=1.4$ and a dimensionless 
    thermal relaxation timescale $\beta \equiv t_c\Omega_k = 0.01$
    (top) and $0.125$ (bottom). The
    perturbation radial wavenumber is $\khat_\mathrm{iso}=10$. The
    real (imaginary) part of $W$ is plotted as the solid (dotted)
    line.
    \label{relax_eigenW_num}}  
\end{figure}

>>>>

\section{Numerical solutions}
In this section we numerically solve the linear stability problem
corresponding to Eq. \ref{lin_mass}---\ref{lin_energy}. 
We relax the thin-disk, low-frequency and nearly-Keplerian
approximations used in the analytical discussion above, and  
use the full expressions of $\rho$, $\Omega$ and
$\kappa$ given in \S\ref{eqm}. We also restore $\hat{g}_c=1$ to
account for the background radial disk structure.

%, so that we may consistently consider vertical domains up to
%$z\sim r$.

@@

% mention mode selection -> focus on fundamental corrugation ->
% simulations eventually dominated by them 
% choice of kx is O(10), since nelson observe kxH \sim 30

\subsection{VSI in a locally isothermal, neutrally stratified disk}\label{vertiso_pertiso} 
% disk model is a nearly-vertically
% isothermal disk with $\Gamma=1.011$, $p=-1.5$, $q=-1$,
% $\epsilon=0.1$. 
We first calculate the VSI in a nearly vertcally isothermal, neutrally
straitified disk ($N_z^2\equiv0$) by setting
$\gamma=\Gamma=1.011$ and $\beta=10^{-3}$. Other disk parameters are 
$(p,q,\epsilon)=(-1.5,-1,0.1)$ and the vertical domain is $\zmax =
5\epsilon r$. This setup is similar to that adopted in
\cite{mcnally14}, who calculated the VSI in vertically isothermal
disks subject to isothermal perturbations ($\Gamma=\gamma=1$).  This
allows us to test our numerical code.

@@

Fig. \ref{iso_eigen_kx} compares the numerically-obtained eigenvalues
for the fundamental mode and that calculated from 
Eq. \ref{sig2_iso} (with $L=1$). There is good agreement for 
$\khat\gtrsim10$. The match worsens for smaller $\khat$ because the 
the low-frequency approximation used to obtain Eq. \ref{sig2_iso} 
is no longer valid, as $|\omega|$ is not small compared to $\Omega_k$
for $\khat\lesssim 10$.

% This means
% that our analytic results should only be applied to perturbations with
% $\khat\gg 1$, i.e. small radial-lengthscales.

@@

Fig. \ref{lowfreq_eigenfunc} compares the fundamental VSI eigenfunction
$W$ for $\khat=1$ and $\khat=20\pi$. The latter value, corresponding
to $k_x= 200\pi/r$, was used by \cite{mcnally14}. We find $W$ is
similar for all $\khat\gtrsim10$ and is proportional to $z$, as
expected for the fundamental mode according to explicit solutions
developed in \S\ref{iso_poly}.

However, for $\khat\lesssim 10$, e.g. $\khat=1$ in 
Fig. \ref{lowfreq_eigenfunc}, the perturbation magnitude near the
boundaries begin to dominate. For even smaller $\khat\lesssim 1$ (not
shown) we find the perturbation is almost entirely concentrated at the
boundaries. Our radially-local approximation is likely invalid for 
these small wavenumber perturbations and we do not consider such modes
further.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/eigenvector_iso_kx1} 
  \includegraphics[width=\linewidth]{figures/eigenvector_iso_kx60}
  \caption{Eigenfunction of the fundamental VSI for perturbation
    wavenumber $\khat=1$ (top) and $\khat=20\pi$ (bottom). The real 
    (imaginary) parts of $W$ are plotted as solid (dotted) lines. 
    \label{lowfreq_eigenfunc}
  }
\end{figure}

In Fig. \ref{lowfreq_eigen} we plot the eigenvalues 
$\sigma = \omega + \ii\nu$ found for $\khat=20\pi$. This plot is
similar to Fig. 3 in \cite{mcnally14}. The fundamental
VSI mode, with one node in $W$, has the smallest $|\sigma|$. The
eigenvalue calculated from Eq. \ref{sig2_iso} with $L=1$ and the 
numerically-obtained value are  
\begin{align*}
  (\omega,\nu) = (-0.3053,0.2606)\epsilon\Omega_k &\quad \text{(from
    Eq. \ref{sig2_iso})},\\
  (\omega,\nu) = (-0.3010,0.2575) \epsilon \Omega_k &\quad \text{(numerical)}.
\end{align*}
This agreement is surpsingly good given the number of approximations
used to obtain Eq. \ref{sig2_iso}, which also imposes a different
boundary condition to that in the numerical calculation. These values
are also close to that found by \cite{mcnally14}. The eigenfrequency
is not sensitive to the vertical boundary condition because the
expression for $\sigma$, given by Eq. \ref{integral_relation1},
involve integrands with $\rho$ as the   weight function, which rapidly
decays away from the midplane.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/eigenvalues_iso}
  \caption{Eigenvalues of perturbations with wavenumber $\khat=20\pi$
    in a neutrally-stratified, nearly-vertically isothermal disk
    ($\gamma=\Gamma=1.011$) with $(p,q,\epsilon)=(-1.5,-1,0.1)$ and
    $\beta=10^{-3}$. \label{lowfreq_eigen} 
  }
\end{figure}

@@

Fig. \ref{bcrit_compare} shows the growth rate $\nu$ of the
fundamental VSI mode as a function of $\beta$ for $\khat=30$. This is
approximately the perturbation wavenumber observed in the fiducial
numerical simulation of \cite{nelson13}. We also plot the growth rate 
obtained by numerically solving the dispersion relation 
Eq. \ref{relax_disp} and the estimate for the upper limit of
the thermal relaxation timescale $\beta_\mathrm{crit}$ given by
Eq. \ref{iso_vsi_cond}. For the present disk parameters
$\beta_\mathrm{crit}=0.125$. We find good agreement between the
numerical and analytical growth rates, as well as the critical
relaxation timescale.

 \begin{figure}
   \includegraphics[width=\linewidth]{figures/bcrit_compare} 
   \caption{Growth rate of the fundamental VSI mode with $\khat=30$ as
     a function of the thermal relaxation timescale $\beta$. The disk
     parameters are $\Gamma=1.011$, $\gamma=1.4$, and 
     $(p,q,\epsilon)=(-1.5,-1,0.05)$. The numerical (analytical) growth
     rates are show as solid (dotted) lines, while the vertical line
     is the maximum thermal timescale $\beta_\mathrm{crit}$ for the
     fundamental mode according to Eq. \ref{iso_vsi_cond}. 
     \label{bcrit_compare}}   
 \end{figure}

% Fig. \ref{relax_growth_num} shows the growth rates of the fundamental  
% mode as a function of $\khat$. This plot is largely consistent with  
% Fig. \ref{relax_disp_fig} except for $\khat\lesssim 1$
% where the low-frequency approximation fails. (In any case, the local model not valid
% for $\khat\ll1$.) For $\khat\gtrsim 10$ introducing thermal
% relaxation rapidly stabilizes the fundamental mode.

% Fig. \ref{relax_growth_num} also confirms our analytical discussion that
% introducing a small but finite cooling time stabilizes the fundamental
% VSI, but there is a preferable wavenumber that maximizes this effect (\S\ref{relax_pert}).
% This occurs at $\khat= O(10)$ for the current disk model.

% \begin{figure}
%    \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm
%    0cm]{figures/compare_eigen_imag_bloop} 
%   % \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
%   % 1cm]{figures/compare_eigen_imag_bloop2} 
%   \caption{Growth rates of the fundamental VSI mode in 
%     nearly-vertically isothermal disks ($\Gamma=1.011$) with
%     $(p,q,\epsilon)=(-1.5,-1,0.05)$, evolved with $\gamma=1.4$ under a
%     range of thermal relaxation timescales   
%     $\beta$. The eigenfrequencies are calculated by numerically 
%     solving the full linear eigenvalue problem. \label{relax_growth_num}}   
% \end{figure}

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/eigenvectorW_beta0d01} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/eigenvectorW_beta0d1} 
  \caption{Numerically-calculated eigenfunction of the fundamental VSI
    mode in a nearly vertically isothermal disk ($\Gamma=1.011$) with
    $(p,q,\epsilon)=(-1.5,-1,0.05)$, evolved with $\gamma=1.4$ and a dimensionless 
    thermal relaxation timescale $\beta \equiv t_c\Omega_k = 0.01$
    (top) and $0.1$ (bottom). The
    perturbation radial wavenumber is $\khat=30$. The
    real (imaginary) part of $W$ is plotted as the solid (dotted)
    line.
    \label{relax_eigenW_num}}  
\end{figure}
